Towards Detecting Stalkerware Through Network Traces

### 2019-03-19 paper/main.tex
Added to methodology, rearranged some things

However, some characteristics of the stalkerware apps themselves may aid us in 
detecting them. For example, similar to malware, some groups of apps appear to 
be repackaged versions of each other. For example, PhoneSpector, 
HighsterMobile, AutoForward, EasySpy and Surepoint Spy all have separate 
websites, but the appearance of each of these sites is nearly identical. Upon 
closer inspection, although multiple companies are listed as the makers of the 
apps (Powerline Group Corp., ILF Mobile Apps, and PhoneSpector LLC), all have 
the same physical address listed. At least one blog that is supposedly run by 
an independent part reviewed apps that coincidentally were made by the same 
company\cite{noauthor_best_nodate}. Other app sites appear to be similar to each other in other ways, 
with many having a large panel with a darkened image background extolling the 
features of the app above two buttons, one for a "live demo" and the other to 
"buy now." This is somewhat circumstantial evidence - it is entirely possible 
that these manufacturers simply chose similar web templates, or copied each 
other's site in an attempt to undercut competition. However, if multiple groups 
of stalkerware apps are in fact made by only a few manufacturers, it may 
facilitate app identification.

After our investigation on these apps, we can confirm that, at least from the few that we have selected, many stalkerware apps are related in some fashion to each other. The most notable being the group of apps stated above, and a single app mentioned before called SpyTomobile. SpyTomobile is an interesting case because it fits the description of a generic stalkerware app so closely that it is the perfect example of spyware on the Play Store. Moreover, this single app has been repackaged multiple times. We have discovered that there are at least 6 copies of the app on the Play store, all with different developers and different names. While this is still circumstantial evidence that most stalkerware often contains copies of each other, we can at least conclude that our forty-five apps constitute a larger percentage than anticipated.

@@

Upon identifying the apps and locating them on the app store or app website, we 
proceeded to install the apps on two phones. One is a (what are our phones). 
After installing and confirming functionality, we began tracking the phone's 
network traffic using Wireshark. For the majority of the apps, a network trace 
of five to ten minutes is enough to detect traffic going to and from the 
stalkerware app as long as the app's tracked functions like text or GPS are 
altered or the tracker app or website has a manual update function. Those that 
do not have such functionality have all been fake stalkerware apps. The network 
captures themselves contain at least one packet that provides easy manual 
detection of the related app.

What causes most stalkerware to send packets? Our captures and testing show that there are three main categories of updating tracker state: manual, timed, and on tracked feature. Most free apps fall into the manual update method, where there is a button on a website or on a phone app that will send a signal packet to the stalkerware app that initalizes the upload. Most paid apps also include a timed auto-update feature, where an update will be sent every few minutes. A very small subset of both advertised that it receives a "real time notification" every time email, texts, calls, or GPS, was changed or added to. Our results show that none of the stalkerware apps were able to send in real time, all the apps advertised as real time simply updated at a more frequent frequency than other apps.

@@

>>>>

However, some characteristics of the stalkerware apps themselves may aid us in 
detecting them. For example, similar to malware, some groups of apps appear to 
be repackaged versions of each other. For example, PhoneSpector, 
HighsterMobile, AutoForward, EasySpy and Surepoint Spy all have separate 
websites, but the appearance of each of these sites is nearly identical. Upon 
closer inspection, although multiple companies are listed as the makers of the 
apps (Powerline Group Corp., ILF Mobile Apps, and PhoneSpector LLC), all have 
the same physical address listed. At least one blog that is supposedly run by 
an independent party reviewed apps that coincidentally were all made by the 
same company\cite{noauthor_best_nodate}. Other app sites appear to be similar 
to each other in other ways, 
with many having a large panel with a darkened image background extolling the 
features of the app above two buttons, one for a "live demo" and the other to 
"buy now." This is somewhat circumstantial evidence - it is entirely possible 
that these manufacturers simply chose similar web templates, or copied each 
other's site in an attempt to undercut competition. We therefore confirmed our 
suspicions by checking if the traffic signatures were also similar. In multiple 
cases, this was true (see section \ref{app_families}). App identification may 
be easier to scale since multiple groups of stalkerware apps are in fact made 
by only a few manufacturers.

\subsection{Experimental Setup}
Out of the 45 apps we originally selected from Dr. Ristenpart's ML-curated list 
of stalkerware, 36 still existed on the app store. We installed these one at a 
time on one or both of two test phones, a Nexus 5X and a Nexus 6P.These models 
were chosen simply because they were readily available. Stalkerware user 
models  fall into two types: either an app is installed on the target phone and 
a login to a web portal is provided for the attacker to view the results, or 
the app comes with an ``attacker'' app and a ``target'' app, and the attacker 
app must be installed on another phone.

After installing an app, we first confirmed that it was functional and capable 
of acting like stalkerware. 
We then began tracking the phone's network traffic using Wireshark. For the 
majority of the apps, a network trace 
of five to ten minutes is enough to detect traffic going to and from the 
stalkerware app as long as the app's tracked functions like text or GPS are 
altered or the tracker app or website has a manual update function. Those that 
do not have such functionality have all been fake stalkerware apps. The network 
captures themselves contain at least one packet that provides easy manual 
detection of the related app.

\subsection{Ethical Considerations}

We created two aliases in order to avoid exposing any personal information to 
an ecosystem that is notorious for mishandling it\cite{ristenpart_ucsd_talk}. 
It is likely that the privacy of victims is 
not a high priority for anyone prepared to monetize abusive relationships, and 
stalkerware manufacturers have leaked victims' information in the past 
\cite{koebler_stalkerware_2018}. We took precautions by deleting all personal 
data that had been on the phones at the start of the project. However, we were 
not quite cautious enough - the app SpyMyFone managed to scrape a list of 
deleted contacts off of the Nexus 5X. As of this moment, we do not know how. We 
have deleted the account and requested that SpyMyFone delete all information 
associated with it from their records, but we have no evidence that they have 
complied. This served as a valuable and sobering lesson for the future.

The original goal of this project was to identify traffic signatures so that 
they could be detected by an IDS (Intrusion Detection System) or a network 
tap in an anonymized fashion. We hoped to use differential privacy techniques 
such as Google's RAPPOR algorithm \cite{erlingsson_rappor:_2014} to collect 
aggregated statistics about stalkerware use, without giving ourselves the 
ability to identify any particular device that had stalkerware installed. Upon 
further investigation before we began the project, this turned out to be 
insufficient protection. We as a computer 
scientists do not have the expertise required to assist victims of abusive 
relationships or domestic violence. We chose not to take our research in this 
direction to avoid causing any unintended harm. We limit ourselves to 
identifying traffic signatures on apps that we installed on personal phones.

Wiretapping the phones was accomplished by setting up our personal computers as 
WiFi hotspots. Since we controlled the passwords required to connect to our 
networks, and Windows alerts the user when a device connects to its mobile 
hotspot, we were able to ensure that we did not tap any traffic other than that 
of the test phones.

@@

What causes most stalkerware to send packets? Our captures and testing show 
that there are three main categories of updating tracker state: manual, timed, 
and on tracked feature. Most free apps fall into the manual update method, 
where there is a button on a website or on a phone app that will send a signal 
packet to the stalkerware app that initializes the upload. Most paid apps also 
include a timed auto-update feature, where an update will be sent every few 
minutes. A very small subset of both advertised that it receives a "real time 
notification" every time email, texts, calls, or GPS, was changed or added to. 
Our results show that none of the stalkerware apps were able to send in real 
time, all the apps advertised as real time simply updated at a more frequent 
frequency than other apps.

@@

\subsection{App ``Families''}
\label{app_families}
After our investigation on these apps, we can confirm that, at least from the 
few that we have selected, many stalkerware apps are related in some fashion to 
each other. The most notable being the group of apps stated above, and a single 
app mentioned before called SpyTomobile. SpyTomobile is an interesting case 
because it fits the description of a generic stalkerware app so closely that it 
is the perfect example of spyware on the Play Store. Moreover, this single app 
has been repackaged multiple times. We have discovered that there are at least 
6 copies of the app on the Play store, all with different developers and 
different names. While this is still circumstantial evidence that most 
stalkerware often contains copies of each other, we can at least conclude that 
our forty-five apps constitute a larger percentage than anticipated.

PUT NOTE HERE ABOUT THE FAMILIES THAT WE DID NOT CONFIRM VIA TRAFFIC, BUT 
INSTEAD VIA LOOKING AT THE WEBSITES!! Otherwise we're fabricating results.

\begin{table*}
\begin{tabular}{p{5cm}p{5cm}p{5cm}}
	\toprule
	Generalized Signature & Unique Apps & Total Apps \\
	\hline
	Domain contains “spy,” “track,” “find” & 6 & 12 \\
	SNI contains “spy,” “track,” “find” & 6 & 6 \\
	User-Agent contains “spy,” “track,” “find” & 1 & 3 \\
	User-Agent contains known app name & 1 & 1 \\
	Domain contains known app name & 3 & 4 \\
	SNI contains known app name & 1 & 1 \\
	SNI contains known app manufacturer & 3 & 3 \\
	Domain contains known app manufacturer & 3 & 3 \\
	\midrule
\end{tabular}
\end{table*}
